Storm Surges

### 2014-07-17 RoughWork/manuscript.tex
Added new figure to text
Added description of surge propagation.

The curvilinear orthogonal numerical grid is divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels are stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10m the grid is stretched gradually to a maximum grid spacing of 27 m at the lowest layer. At the bottom boundary, partial $z$-levels are utilized in order to limit large changes in bathymetry across grid cells \citep{madec2008nemo}.

@@

\subsection{Boundary conditions and sub-grid scale processes}
%lateral and bottom friction, open boundaries, eta surface
The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents, temperature and salinity climatologies, and the sea surface height anomaly. The details of these forcing conditions are described below. At coastal boundaries, the partial slip boundary condition, an approximation to no slip, is used. The partial slip boundary condition allows one to include the frictional effects of lateral boundaries without the restrictive resolution required to represent the lateral boundary layer under no slip conditions. A lateral eddy viscosity of 20 m $^2$ s $^{-1}$ parametrizes horizontal friction and a lateral eddy diffusivity of 20.5 m $^2$ s $^{-1}$ is used. At the ocean surface, meteorological conditions and fresh water input from rivers are included and described in more detail below. Bottom friction is represented by a quadratic law for the bottom momentum flux with drag coefficient $C_D = 5\times 10^{-3}$. Vertical turbulence and mixing is calculated through the $k-\epsilon$ configuration of the generic length scale (GLS) turbulence closure \citep{umlauf2003generic} with background vertical eddy viscosity and diffusivity set to $1\times10^{-4}$ and $1\times10^{-5}$ m$^2$ s$^{-1}$ respectively. Details on the NEMO implementation of the partial slip lateral boundary condition, quadratic bottom friction law, and GLS turbulence closure scheme are provided by \citet{madec2008nemo}.

@@

The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?
The tidal forcing and sea surface height was used in the barotropic velocity forcing which used the NEMO Flather scheme \citep{flather1994storm, madec2008nemo}.
The baroclinic velocities at the boundary were set to be equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain.

@@

Three significiant surge or wind events have been considered in detail, as well as an array of hindcasts to asses model ability in a statistical measure. First, a study of a large surge event on Feb 4, 2006 is presented. This event caused siginificant damage and flooding in Boundary Bay, a community south of Richmond. (citation) and has been chosen as a case study to demonstrate the model's skill in reproducing a storm surge and to quantitatively assess forcing conditions that are important in storm surge hindcasting. 
%% Trying to decide Dec 15, 2005 or Nov 2009. I think it's more sense to do Nov 2009 - strong winds but no surge. 
Second, an extreme wind event which resulted in property damage in Vancouver and surrounding areas is discussed. This event occured on Dec 15, 2006 but did not result in significant flooding such as that which occured on Feb 4, 2006 because the storm did not occur at high tide. This case was choosen to identify the model's capability in handling large wind events that do not result in high water levels. Third, a case with high resolution atmospheric forcing for a recent surge on Dec 17, 2012 is discussed. Lastly, these cases presented in detail will be combined with $x$ additional runs for statistical evaluation of model performance.

@@

Finally, an assessment of the importance of including tide-surge interactions is presented next. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}. Recall, the modelled residual is defined as the difference between the sea surface height in a simulation with all forcing and a simulaltion with tides amd rivers only. A comparison between the modelled residual and a surge only run can determine the significance of the tide-surge interaction. This figure shows there is almost no difference between the simulations at all locations, rendering the tide-sure interaction negligible. In operational settings, running a surge only simulation and adding the predicted tides a posteriori could be more computationally efficient with only a small decrease in the accuracy of the prediction. 
% I'm not sure if remocing a background state is really appropriate here. There could be some differences in the background due to the removal of the tides. This also suggests it is more appropriate to use the atmospheric pressure to remove the background...

@@

\subsection{Strong wind event: Dec 15, 2006}% or Nov 2009

>>>>

The curvilinear orthogonal numerical grid is divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels are stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10m the grid is stretched gradually to a maximum grid spacing of 27 m at the lowest layer. At the bottom boundary, partial $z$-levels are utilized in order to limit large changes in bathymetry across grid cells \citep{madec2012nemo}.

@@

\subsection{Boundary conditions and sub-grid scale processes}
%lateral and bottom friction, open boundaries, eta surface
The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents, temperature and salinity climatologies, and the sea surface height anomaly. The details of these forcing conditions are described below. At coastal boundaries, the partial slip boundary condition, an approximation to no slip, is used. The partial slip boundary condition allows one to include the frictional effects of lateral boundaries without the restrictive resolution required to represent the lateral boundary layer under no slip conditions. A lateral eddy viscosity of 20 m $^2$ s $^{-1}$ parametrizes horizontal friction and a lateral eddy diffusivity of 20.5 m $^2$ s $^{-1}$ is used. At the ocean surface, meteorological conditions and fresh water input from rivers are included and described in more detail below. Bottom friction is represented by a quadratic law for the bottom momentum flux with drag coefficient $C_D = 5\times 10^{-3}$. Vertical turbulence and mixing is calculated through the $k-\epsilon$ configuration of the generic length scale (GLS) turbulence closure \citep{umlauf2003generic} with background vertical eddy viscosity and diffusivity set to $1\times10^{-4}$ and $1\times10^{-5}$ m$^2$ s$^{-1}$ respectively. Details on the NEMO implementation of the partial slip lateral boundary condition, quadratic bottom friction law, and GLS turbulence closure scheme are provided by \citet{madec2012nemo}.

@@

The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?
The tidal forcing and sea surface height was used in the barotropic velocity forcing which used the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}.
The baroclinic velocities at the boundary were set to be equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain.

@@

Several significiant surge or wind events have been considered in detail, as well as an array of hindcasts to asses model ability in a statistical measure. These cases have been used to asses the model's skill in reproducing water level height and surge amplitude and timing during storm events. In addition, the contribution to the surge of several forcing conditions, such as atmospheric forcing and open boundary forcing, has been assessed. We have also examined the spatial variabilty of the surge as it propagates through the domain and have considered a case with high-resolution atmospheric forcing.

%First, a study of a large surge event on Feb 4, 2006 is presented. This event caused siginificant damage and flooding in Boundary Bay, a community south of Richmond, (citation) and has been chosen as a case study to demonstrate the model's skill in reproducing a storm surge and to quantitatively assess forcing conditions that are important in storm surge hindcasting. Second, an extreme wind event on Dec 15, 2006 is examined. This event produced winds over (x) (citation) but did not result in significant flooding because the storm did not occur at high tide, however, it did result in siginificant property damage due to high winds and waves. This event has been used to understand the spatial and temporal variability of the surge propagation. Next, a large wind event on Nov 18, 2009 that did not result in a siginificant surge or flooding is presented. This case was choosen to identify the model's capability in handling large wind events that do not result in high water levels. Next, a case with high resolution atmospheric forcing for a recent surge on Dec 17, 2012 is discussed. Lastly, these cases presented in detail will be combined with $x$ additional runs for statistical evaluation of model performance.

@@

Finally, an assessment of the importance of including tide-surge interactions is presented next. A simulation of the surge propagation without including any tidal forcing is compared with the modelled residual in Figure \ref{fig:tidesurge}. Recall, the modelled residual is defined as the difference between the sea surface height in a simulation with all forcing and a simulaltion with tides amd rivers only. A comparison between the modelled residual and a surge only run can determine the significance of the tide-surge interaction. This figure shows there is almost no difference between the simulations at all locations, rendering the tide-sure interaction negligible. In operational settings, running a surge only simulation and adding the predicted tides a posteriori could be more computationally efficient with only a small decrease in the accuracy of the prediction. 
% I'm not sure if removing a background state is really appropriate here. There could be some differences in the background states because of the tides(ie where fresh water moves to, etc). I also think it is more appropriate to use the atmospheric pressure to remove the background state...

@@

\subsection{Spatial Extent of Surge: Dec 15, 2006}
On Dec 15, 2006, a major storm with wind gusts over 120 km/hr caused significant damage to many coastal communities on the Pacific Northwest. In British Columbia, over 250,000 residents were left without power due to downed trees and hydro lines. Although the storm was very strong and the observed residual at Point Atkinson reached 85 cm, this event did not coincide with high tide and so no siginificant flooding occurred in the Vancouver area\citep{forseth2006adaptation}. This event has been used to examine the propgation of the surge throughout the domain. First, an assessment of the accuracy of the hindcast is provided.

The evolution and spatial distribution of the surge is highlighted in Figure \ref{fig:spatial} where the modelled residual is plotted every two hours beginning on Dec 14 at 18:30. The surge enters the system from the Strait of Juna de Fuca and Johnstone Strait, peaking between Dec 15 at 6:30 and Dec 15 at 10:30 and then begins to subside throughout the entire domain. Notably, the Strait of Georgia experiences the largest surge values, upwards of 70 cm, whereas the Juan de Fuca region experiences a maximum surge around 55 cm. Additionally, the coastal areas on the mainland side of the Strait of Georgia experience slightly higher surges than the Vancouver Island side, likley due to the southerly wind in the atmosheric model pushing water against the mainland coast.

\begin{figure}
\centering
\includegraphics[scale=0.6]{Figures/dec2006_spatial.pdf}
\caption{Propagation of the storm surge on Dec 14/15 2006. Modelled residual every 2 hours starting on Dec 14 at 18:30. Time advances from top left to bottom right. Note that the images are aligned with the numerical grid which is rotated $29^{\circ}$ counter-clockwise of North. }
\label{fig:spatial}
\end{figure}

\subsection{Strong Wind Event: Nov 18, 2009}
